---
title: "hwk_1"
author: "Holden Jones"
date: "2025-01-13"
output: html_document
---

setup
```{r}
library(tidyverse)
bird_data <- read_csv('data/foraging.height.edit.csv')
```

# 1. 
Start with some data exploration. 

How many observations are there of each bird species in each experimental treatment? 
```{r}
bird_data %>%
  group_by(SPECIES, Rat_Removal) %>%
  summarize(count = n())
```

Make a boxplot or violinplot to show the distribution of foraging heights of each bird species in each experimental treatment. 
```{r}
ggplot(bird_data, aes(Rat_Removal, foraging.ht.m, 
                 fill = SPECIES)) +
  geom_violin()
```

What have you learned so far?
- Lots of variation in the number of observations for each species, but they are closely split 50/50
  across treated / untreated sites.
- Seems to be more variation in foraging height for most species in the treated sites. They forage at
  lower heights in the treated sites.

# 2.

Now make a plot showing the mean foraging height of each bird species in each
treatment, and include error bars displaying +/- one standard error of the mean.
```{r}
ggplot(bird_data, aes(Rat_Removal, foraging.ht.m, 
                 fill = SPECIES)) +
  geom_boxplot()
```

What is the meaning of the standard error of the mean? How does this plot differ
from the plot in #1? 
- Standard Error of the Mean incorporates variation within a sample and sample size to show how likely the observed sample mean is to be within the true population mean
- This plot incorporates standard error with error bars which contain the true mean with 95% confidence

# 3.
Fit a linear model that tests whether the bird species differ in foraging height,
whether the rat removal treatment affects foraging height, and whether the effect of
the rat removal differs between the bird species. 
```{r}
height_model <- lm(foraging.ht.m ~ SPECIES + Rat_Removal + Rat_Removal*SPECIES , 
            data = bird_data)

summary(height_model)
```

Make plots to assess the distribution of the residuals. 
```{r}
par(mfrow = c(2, 2))
plot(height_model)
```

Report F-tests for the terms of the model. 

Create an effects plot displaying the fitted effects. 

How do you interpret the results? What are the magnitudes of the effects? 






