---
title: "hwk_3"
author: "Holden Jones"
date: "2025-02-03"
output: html_document
---

```{r}
library(tidyverse)
library(effects)
library(DHARMa)

fish_data <- read_csv('data/fish_data.csv')
```

# 1.
(1) Let’s look at some basic summary statistics. In total there are 209 fish 
species encountered and counted in these surveys. Identify the top twelve species 
in terms of these three statistics: (1) mean abundance, (2) maximum abundance, 
(3) standard deviation of abundance. In other words, which species are most 
abundant on average, which reach the highest abundances, and which have the most 
variable abundances?
```{r}
# top 12 highest mean abundance
fish_data %>%
  group_by(taxonname) %>%
  summarize(mean_abun = mean(count, na.rm = TRUE)) %>%
  slice_max(order_by = mean_abun, n = 12)

# top 12 highest max abundance
fish_data %>%
  group_by(taxonname) %>%
  summarize(max_abun = max(count, na.rm = TRUE)) %>%
  slice_max(order_by = max_abun, n = 12)

# top 12 highest sd abundance
fish_data %>%
  group_by(taxonname) %>%
  summarize(sd_abun = sd(count, na.rm = TRUE)) %>%
  slice_max(order_by = sd_abun, n = 12)
```

# 2. 
(2) Now let’s visualize how the abundances of the most common species vary with 
depth, to get a sense for whether species have different depth niches. Using the 
top twelve species based on mean abundance, plot a scatterplot of count vs. depth 
(the column is named ‘depth’). Add a smoother to help visualize mean count vs. 
depth, put all twelve species on a single plot (with twelve panels), and make 
sure each panel is titled using the species name. Furthermore, do not arrange the 
panels in alphabetical order — instead, arrange them in order of mean abundance, 
so that the most abundant species is in the top left panel, and remaining panels 
are in order of mean abundance. The patterns in abundance will be easier to see 
if you transform the counts or the y-axis, for example with a square root, 
because the counts are very skewed.

start by making a vector of the top 12 sp in abundance - subset df to just these
```{r}
top_taxa <- fish_data %>%
  group_by(taxonname) %>%
  summarize(mean_abun = mean(count, na.rm = TRUE)) %>%
  slice_max(order_by = mean_abun, n = 12) %>%
  pull(taxonname)
top_taxa # ranked top 12 species - chromis vanderbelti first

# filter df taxonname ranked by top_taxa
filter_fish <- fish_data %>%
  filter(taxonname %in% top_taxa) %>%
  mutate(taxonname = factor(taxonname, levels = top_taxa))
```

for loop plot
```{r}
par(mfrow = c(3, 4))  

for (i in top_taxa) { # loop over the top 12 taxa
  plot_data <- filter_fish %>% filter(taxonname == i) # filter for i taxa
  
    scatter.smooth(plot_data$depth, log1p(plot_data$count), # log1p solved 0 problem
               span = 1,  
               main = i, 
               xlab = "Depth", 
               ylab = "Log(Count + 1)", 
               col = "blue")
}
```

ggplot
```{r}
count_depth_plot <- filter_fish %>%
                    ggplot(aes(x = depth, y = log1p(count))) +
                      geom_point() +
                      geom_smooth() +
                      facet_wrap(~ taxonname)
count_depth_plot
```

What are your conclusions from this visual inspection of the data? Make sure that 
the patterns in the plots are visible. If they are not, you may need to adjust 
the figure dimensions in Markdown.

- Abundance peaks at intermediate depth for the two most abundant species. Each
  species responds differently to depth. For example the 3rd and 5th most 
  abundant species decrease in abundance as depth increases. C. hanui actually 
  increases in abundance at higher depths.
- This suggests a quadratic model might be appropriate, esp. for some species
  like C. vanderbilti and C. agillis
- P. kallopteris had extremely low frequency of being detected, so I'd probably
  remove this from analysis. We don't know enough about it to say how it 
  responds to depth yet.

# 3.
(3) Finally, let’s focus on just the top 5 most abundant species. Make a new plot 
that shows abundance vs. depth for the top 5 species, including smoothers, but 
this time put all of the species on the same scatterplot and distinguish them 
with different colors. This time you don’t need to use two different approaches 
to make the plot — one approach will suffice.

subset by only the top 5 taxa
```{r}
top_five <- fish_data %>%
  group_by(taxonname) %>%
  summarize(mean_abun = mean(count, na.rm = TRUE)) %>%
  slice_max(order_by = mean_abun, n = 5) %>%
  pull(taxonname)
top_five # ranked top 5 species - chromis vanderbelti first

# filter df taxonname ranked by top_five
filter_fish <- fish_data %>%
  filter(taxonname %in% top_five) %>%
  mutate(taxonname = factor(taxonname, levels = top_five))
```

plot top 5
```{r}
count_depth_plot_5 <- filter_fish %>%
                    ggplot(aes(x = depth, y = log1p(count),
                               color = taxonname)) +
                      geom_point() +
                      geom_smooth()
count_depth_plot_5
```

What is your interpretation of this plot?

- The top 5 most abundant species all respond differently to depth!
- C. vanderbilti and C. agillis peak at intermediate depths
- C. strigosus appears to have bimodal peaks
- A. nigrofuscus and Z. flavescens both appear to decrease in abundace with depth

Fit a linear model that tests whether these 5 species have different depth niches. 
```{r}
depth_model <- lm(log1p(count) ~ depth + taxonname + 
                   depth*taxonname, 
            data = filter_fish)
summary(depth_model)
```

residual diagnostics and effects plots
```{r}
par(mfrow = c(2, 2))
plot(depth_model)

plot(allEffects(depth_model))
```

confirm using dHARMA
```{r}
simulationOutput <- simulateResiduals(fittedModel = depth_model)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

Based on residual diagnostic plots of this model, and a plot of the fitted 
effects, do you think a linear model is a good approach for testing this question? 
If not, why not?

- No, I do not think a linear model is a good approach for asking if there is a
  reaction between species and depth for abundance
- The residual diagnostic plots look horrible! Residuals clearly are not evenly
  distributed across all depth values. So a linear model is not accurately 
  capturing the effect of different depths on abundance for different species
- Seems like we need to use a quadratic model...
