---
title: "hwk_7"
author: "Holden Jones"
date: "2025-04-11"
output: html_document
---

```{r}
library(tidyverse)
library(glmmTMB)
library(car)
library(DHARMa)
library(ggeffects)
library(performance)

bugs <- read_csv('data/collembola.csv')

# factor temp.factor and predators for plotting
bugs$temp.factor <- factor(bugs$temp.factor)
bugs$predators<- factor(bugs$predators)
```

focus on springtail abundance and how affected by temp and mite predation

#1.

huge variation in total.prey
- lots of zeros
```{r}
hist(bugs$total.prey)
```


fit standard model for counts (Poisson), but assess if overdispersed
- is abundance of Collembola affected by temp, predator presence, and interaction
- fit with glmmTMB()
```{r}
model_full <- glmmTMB(
  total.prey ~ temp.factor + predators + temp.factor*predators,
  family = poisson,
  data = bugs
)
summary(model_full) # with Poisson almost every term is sig
```

plot fitted effects and perform likelihood ratio tests on terms
```{r}
Anova(model_full) # all have sig effect

sim_res_model <- simulateResiduals(fittedModel = model_full)
plot(sim_res_model) # interesting, significant deviation suggesting overdispersion
check_overdispersion(model_full) # overdispersion detected

plot(ggpredict(model_full, terms = c("temp.factor", "predators")))
```

let's try negative binomial to see if this better addresses residual deviation
```{r}
model_nb <- glmmTMB(
  total.prey ~ temp.factor + predators + temp.factor*predators,
  family = nbinom2,
  data = bugs
)
summary(model_nb) # with neg binom none of these variables are sig - huh
```

plot fitted effects and perform likelihood ratio tests on terms
```{r}
Anova(model_nb) # no sig effect

sim_res_model <- simulateResiduals(fittedModel = model_nb)
plot(sim_res_model) # significant deviation suggesting overdispersion
check_overdispersion(model_nb) # interesting, not overdispersed!

plot(ggpredict(model_nb, terms = c("temp.factor", "predators")))
```

neither negative binomial or Poisson appear to fit data well (sig deviation)
- but, Poisson has overdispersion, negative binomial does not
- suggests that will need to do zero-inflation instead
- but move forward with using anova() for marginal tests with negative binomial model

perform marginal tests compare pairs of models using anova()
- for each term to test, compare a model with this term to model without this term
```{r}
# no temp.factor
model_no_temp <- glmmTMB(
  total.prey ~ predators + temp.factor*predators,
  family = nbinom2,
  data = bugs
)
summary(model_no_temp) # nothing sig

# no predators
model_no_pred <- glmmTMB(
  total.prey ~ temp.factor + temp.factor*predators,
  family = nbinom2,
  data = bugs
)
summary(model_no_pred) # nothing sig

# no interaction
model_no_int <- glmmTMB(
  total.prey ~ predators + temp.factor,
  family = nbinom2,
  data = bugs
)
summary(model_no_int) # nothing sig
```

run comparisons between candidate models using anova
```{r}
# not sure how to address interaction, seems important to compare against full
anova(model_full, model_no_temp) # wow! removing temp sig increases model fit compared to full
anova(model_full, model_no_pred) # wow! removing pred sig increases model fit compared to full

anova(model_no_temp, model_no_pred) # neither pred nor temp have any impact on model fit
anova(model_no_temp, model_no_int) # again, minimal to no impact of temp
anova(model_no_pred, model_no_int) # again, minimal to no impact of pred
```

how do we interpret results so far?
- it doesn't seem like any of these predictors (pred, temp, interaction) have
  any sort of sig effect on prey count. 
- removing them greatly improves model fit compared to the full model
- but, tread lightly b/c there are a lot of variables in this dataset, which
  suggests that zero-inflation could be at play. Also based on the experimental
  design, it seems like there would be many instances where the prey ate all of the
  predators. 
- I used the negative binomial over the Poisson distribution for the full model,
  but I'm not confident about this being the best fit. Even though the negative
  binomial didn't contain overdispersion (the Poisson did), it still contained
  significant deviation suggesting the model fit can be improved with a different
  distribution.


#2.

use glmmTMB to fit zero-inflated count models
- default will be Poisson, but depending on how goes could use nbinom or quassi
```{r}
# zero-inflated model using Poisson for count data
model_zin_pois <- glmmTMB(total.prey ~ temp.factor + predators + temp.factor*predators, 
                          data = bugs, 
                          family = poisson, 
                          ziformula = ~ 1)
summary(model_zin_pois) # most terms sig, AIC 6478.2

# zero-inflated model using negative binomial for count data
model_zin_nbinom <- glmmTMB(total.prey ~ temp.factor + predators + temp.factor*predators, 
                            data = bugs, 
                            family = nbinom2, 
                            ziformula = ~ 1)
summary(model_zin_nbinom) # nothing sig, AIC 673.6
```

AIC for model_zin_nbinom is much lower, assess model fit to confirm not overdispersed
```{r}
Anova(model_zin_nbinom) # temp.factor sig effect, almost sig interaction

sim_res_model <- simulateResiduals(fittedModel = model_zin_nbinom)
plot(sim_res_model) # no deviation or overdispersion detected!
check_overdispersion(model_zin_nbinom) # ot overdispersed!

plot(ggpredict(model_zin_nbinom, terms = c("temp.factor", "predators")))
```

appears zero-inflated negative binomial model is the best, run anova with others to confirm
- from ggpredict
    - low prey survival at low temp,
    - either predator makes a big negative impact on survival at med temp
    - at high temp, seems like HA can't survive, but prey can (b/c prey survives w/HA)
    - HM appears to still be effective predator at high temp, more resilient than HA

AIC to perform model selection against nbinom from above
- Yes, model_zin_nbinom has the lowest AIC by far, both nbinom models better than poisson
```{r}
# use model_nb and model_full to compare against models with zero-inflation
anova(model_full, model_nb, model_zin_pois, model_zin_nbinom)
```

using best model, marginal likelihood ratio tests on predictors
- stepwise remove terms
```{r}
# model_zin_nbinom from above
summary(model_zin_nbinom) # AIC 673.6

# model_no_temp
model_no_temp <- glmmTMB(total.prey ~ predators + temp.factor*predators, 
                           data = bugs, 
                            family = nbinom2, 
                            ziformula = ~ 1)
summary(model_no_temp) # nothing sig, AIC 673.6, temp has no impact 

# model_no_pred
model_no_pred <- glmmTMB(total.prey ~ temp.factor + temp.factor*predators, 
                            data = bugs, 
                            family = nbinom2, 
                            ziformula = ~ 1)
summary(model_no_pred) # nothing sig, AIC 673.6, pred has no impact 

# model_no_int
model_no_int <- glmmTMB(total.prey ~ predators + temp.factor, 
                            data = bugs, 
                            family = nbinom2, 
                            ziformula = ~ 1)
summary(model_no_int) # AIC 671.9, removing interaction improves model fit
# sig effect of temp, higher survival at higher temperatures
```

anova to compare these candidate models against full zero-inflated negative binomial
- the model without the interaction fits the best (lowest AIC)
```{r}
anova(model_zin_nbinom, model_no_temp, model_no_pred, model_no_int)
```

how have results changed 
- the zero-inflated negative binomial fits much better than non zero-inflated models
- the best fitting candidate model did not contain the interaction term
- in this model we saw a significantly higher survival of prey in the medium and high
  temperature treatments compared to the low temperature treatment
  
  
#3.

doesn't seem like big difference b/ween predator treatments

create new column, predator 0 or 1 (this will become a binomial)
```{r}
bugs <- bugs %>%
  mutate("predator_binary" = ifelse(predators == "none",0 , 1))
```

model_no_int was the highest performing, use this with new predator_binary
```{r}
# model_no_int_2
model_no_int_2 <- glmmTMB(total.prey ~ predator_binary + temp.factor, 
                            data = bugs, 
                            family = nbinom2, 
                            ziformula = ~ 1)
summary(model_no_int_2) # AIC 668.0
# predators have a sig negative effect on survival
# temp has a sig positive effect on survival
```

```{r}
Anova(model_no_int_2) # temp.factor sig effect, predator_binary sig effect

sim_res_model <- simulateResiduals(fittedModel = model_no_int_2)
plot(sim_res_model) # deviation detected!
check_overdispersion(model_no_int_2) # not overdispersed!

plot(ggpredict(model_no_int_2, terms = c("temp.factor", "predator_binary")))
```

we now see that both the presence of predators, and the temperature have significant
effects on survival
- these patterns are similar to what I saw previously, but the effect of each
  variable is a bit easier to see now.
  
  
#4.

zeros could vary across experimental treatments
- this makes sense, maybe at lower temperatures more things die?
  
use an appropriate zero-inflated model and allow extra zeros to vary 
  w/ temp, pred, interaction
```{r}
# use model_no_int_2 as base, this was the best fitting model (lowest AIC)
model_no_int_3 <- glmmTMB(total.prey ~ temp.factor + predator_binary, 
                            data = bugs, 
                            family = nbinom2, 
                            ziformula = ~ temp.factor*predator_binary)
summary(model_no_int_3)

model_no_int_4 <- glmmTMB(total.prey ~ temp.factor + predator_binary, 
                            data = bugs, 
                            family = nbinom2, 
                            ziformula = ~ temp.factor) 
summary(model_no_int_4) # best fit, AIC 654

model_no_int_5 <- glmmTMB(total.prey ~ temp.factor + predator_binary, 
                            data = bugs, 
                            family = nbinom2, 
                            ziformula = ~ predator_binary) 
summary(model_no_int_5)
```

likelihood ratio tests between these models
- allowing zero inflation to change based on temperature has the best fit
```{r}
anova(model_no_int_3, model_no_int_4, model_no_int_5)
# lowest AIC is model_no_int_4

# run diagnostics on model_no_int_4
Anova(model_no_int_4) # temp.factor sig effect, predator_binary sig effect

sim_res_model <- simulateResiduals(fittedModel = model_no_int_4)
plot(sim_res_model) # no deviation detected!
check_overdispersion(model_no_int_4) # not overdispersed!

plot(ggpredict(model_no_int_4, terms = c("temp.factor", "predator_binary")))
```

the best fitting model includes zero-inflation varying by temperature
- at lower temperatures prey are more likely to die, so greater zeros at low temp
- this model fits better than all other iterations
- in this best fitting model, temperature has a significant, positive effect on
  survival, and predator_binary has a negative, significant effect on survival
- importantly, this model is no longer overdispersed or containing deviation
- so this model fits significantly better than all other attempts