---
title: "hwk_2"
author: "Holden Jones"
date: "2025-01-28"
output: html_document
---

factorially implemented experimental treatments (N:P ratio, light, nutrient level,
grazer presence, grazer predator presence)
- only use balanced N:P treatment, include both light levels but not look at effects
  of light
```{r}
library(tidyverse)
library(effects)
library(emmeans)

data <- read_csv('data/leibold_mesocosm_data_subset.csv')
```

# 1. 
Create a linear model that tests whether richness is explained by nutrient level 
and/or food web treatment, and whether the effect of food web treatment differs 
between nutrient levels, while also accounting for the blocked structure of the 
experiment. 
```{r}
rich_model <- lm(Phyto_Chao1 ~ NutrientLevel + FoodWeb + 
                   NutrientLevel*FoodWeb + Block, 
            data = data)
summary(rich_model)
```

#2. 
Why is it important to account for the blocks? 
- Eventually will want to include block as a random effect, right?
- Important to account for b/c different blocks could have different light
  availability, freshwater sources.. etc. lots of different things depending on
  how the study was performed

# 3.
Analyze the results of the linear model as we did in Homework 1. What is your 
interpretation of these results so far?
```{r}
par(mfrow = c(2, 2))
plot(rich_model)
# residual plots look good - not concerned about scaling data

# anova
anova(rich_model)

# effects plot
par(mfrow = c(2, 2))
plot(allEffects(rich_model))

ggplot(data, aes(x = Block, y = Phyto_Chao1)) +
  geom_boxplot()
```

treat block as a factor
```{r}
# seems like need to treat Block as factor
data$Block <- as.factor(data$Block)

ggplot(data, aes(x = Block, y = Phyto_Chao1)) +
  geom_boxplot()

rich_model <- lm(Phyto_Chao1 ~ NutrientLevel + FoodWeb + 
                   NutrientLevel*FoodWeb + Block, 
            data = data)
summary(rich_model)

par(mfrow = c(2, 2))
plot(rich_model)
# residual plots look good - not concerned about scaling data

# anova
anova(rich_model)

# effects plots
par(mfrow = c(2, 2))
plot(allEffects(rich_model))
```

- treating block as a factor
- food web has a significant effect on richness, with the algae and grazers treatment
  having significantly higher richness than just algae or both with predators
- high nutrient level also has a significantly higher richness than low nutrient

Now we will use contrasts to test some hypotheses, and we will imagine that we 
formulated these hypotheses a priori (as opposed to formulating them after we 
looked at the data). We would like to know specifically whether the grazer 
treatment (G) has greater richness than the algae treament (A), and whether the 
effect of grazers differs between high and low nutrient levels (we think the 
effect of grazers on coexistence may be greater at high nutrient loading).

# 4.
Use emmeans to calculate the estimated marginal means of each combination of 
nutrient level and food web treatment (i.e., H + A, H + G, H + P, L + A, L + G,
L + P). 
```{r}
nut_food_emmeans = emmeans(rich_model, specs = trt.vs.ctrl ~ NutrientLevel*FoodWeb)
nut_food_emmeans
```

- at a 0.05 p-value threshold, only sig difference is LG vs. HA, which probably
  isn't too meaningful (low grazer vs high algae?)
- HG is very likely more than HA (p-value 0.11)

# 5.
Now define contrasts to test (1) whether G is different from A in the H 
treatment, (2) whether G is different from A in the L treatment
```{r}
# first here's all of the contrasts plotted:
plot(nut_food_emmeans$contrasts)

# take a look at order of levels in nut_food_emmeans 
nut_food_emmeans

custom_contrasts <- contrast(nut_food_emmeans, method = list(
         "H_G_vs_H_A" = c(-1, 0, 1, 0, 0, 0),
         "L_G_vs_L_A" = c(0, -1, 0, 1, 0, 0)
         )
         )  
custom_contrasts

# provides the same p-values as the non-adjusted contrast:
nut_food_emmeans_no_adjust = emmeans(rich_model, specs = trt.vs.ctrl ~ NutrientLevel*FoodWeb, adjust = "none")
nut_food_emmeans_no_adjust
```

- G has significantly higher richness than A in both the H and L treatments, by a comparable 
  amount (between 6.11-8.82 species)

#6.
(3) whether G is different from A when averaging over the L and H treatments.
```{r}
# want a new emmeans without the interaction with nutrient level
nut_emmeans = emmeans(rich_model, specs = trt.vs.ctrl ~ FoodWeb)
nut_emmeans
# results may be misleading? hmm...

# I'm not sure if we also need to make a new model, warning suggests so...
simple_model <- lm(Phyto_Chao1 ~ FoodWeb + NutrientLevel + Block, 
            data = data)
summary(simple_model)

nut_emmeans = emmeans(simple_model, specs = trt.vs.ctrl ~ FoodWeb)
nut_emmeans

# one custom contrast
custom_contrast <- contrast(nut_emmeans, method = list(
         "G_vs_A" = c(-1, 1, 0)
         )
         )  
custom_contrast
```

- yes, G has a significantly higher species richness than A when averaging
  across nutrient level and block treatments

#7.
Finally, (4) define an interaction contrast that tests whether the difference 
between G and A is itself different between L and H treatments. I.e., is (G – A) 
for the high nutrient treatments greater or smaller than (G – A) for the 
low nutrient treatments?
```{r}
nut_food_emmeans

interaction_contrast <- contrast(nut_food_emmeans, 
         list("HG_min_HA" = c(-1, 0, 1, 0, 0, 0),
              "LG_min_LA" = c(0, -1, 0, 1, 0, 0)
              ))
interaction_contrast
```

- it looks like the difference between G and A is larger in the low nutrient
  than the high nutrient treatment, although this difference is pretty small (.7).
  It is a significant difference in both treatments.

#8.
How do you interpret these results?

- The G treatments appear to have higher species richness than the A treatments,
  this difference is slightly larger in the low nutrient than high nutrient 
  mesocosms.

#9.
Now, repeat the same set of 4 contrasts, but this time ask whether the P treatment 
is different from the G treatment. We are interested in these contrasts a priori 
because we think that adding a predator of the grazers may mean that grazers have 
a weaker effect on phytoplankton diversity, and that this trophic cascade may be 
more important under high nutrient supply. 

Now define contrasts to test (1) whether G is different from P in the H 
treatment, (2) whether G is different from P in the L treatment.
```{r}
# first here's all of the contrasts plotted:
plot(nut_food_emmeans$contrasts)

# take a look at order of levels in nut_food_emmeans 
nut_food_emmeans

custom_contrasts <- contrast(nut_food_emmeans, method = list(
         "H_G_vs_H_P" = c(0, 0, 1, 0, -1, 0),
         "L_G_vs_L_P" = c(0, 0, 0, 1, 0, -1)
         )
         )  
custom_contrasts

# provides the same p-values as the non-adjusted contrast:
nut_food_emmeans_no_adjust = emmeans(rich_model, specs = trt.vs.ctrl ~ NutrientLevel*FoodWeb, adjust = "none")
nut_food_emmeans_no_adjust
```

- G has significantly higher richness than P in the H treatment, but not in the L
  treatment

(3) whether G is different from P when averaging over the L and H treatments.
```{r}
# want a new emmeans without the interaction with nutrient level
nut_emmeans = emmeans(rich_model, specs = trt.vs.ctrl ~ FoodWeb)
nut_emmeans
# results may be misleading? hmm...

# I'm not sure if we also need to make a new model, warning suggests so...
simple_model <- lm(Phyto_Chao1 ~ FoodWeb + NutrientLevel + Block, 
            data = data)
summary(simple_model)

nut_emmeans = emmeans(simple_model, specs = trt.vs.ctrl ~ FoodWeb)
nut_emmeans

# one custom contrast
custom_contrast <- contrast(nut_emmeans, method = list(
         "G_vs_P" = c(0, 1, -1)
         )
         )  
custom_contrast
```

- yes, G has a significantly higher species richness than P when averaging
  across nutrient level and block treatments

Finally, (4) define an interaction contrast that tests whether the difference 
between G and P is itself different between L and H treatments. I.e., is (G – P) 
for the high nutrient treatments greater or smaller than (G – P) for the 
low nutrient treatments?
```{r}
nut_food_emmeans

interaction_contrast <- contrast(nut_food_emmeans, 
         list("HG_min_HP" = c(0, 0, 1, 0, -1, 0),
              "LG_min_LP" = c(0, 0, 0, 1, 0, -1)
              ))
interaction_contrast
```

#10.
How do you interpret the results?

- The difference between G and P is only significant in the high nutrient treatment,
  where G has 6.51 more species than P. In the low treatment the difference
  between G and P is insignificant.




