---
title: "hwk_12"
author: "Holden Jones"
date: "2025-05-12"
output: html_document
---


```{r}
library(tidyverse)
library(lme4)
library(sjPlot)
library(MuMIn)
library(DHARMa)
library(glmmTMB)
library(lmerTest)

island <- read_csv('data/cabral_island_data.csv') %>%
  select(!'...1')

arch <- read_csv('data/cabral_arch_data.csv') %>%
  select(!'...1') %>%
  slice(1:23)
```

data naturally organized into groups (archipelagoes), account for this structure
- non-independence in the model
- need archipelago random effect to avoid pseudoreplicating predictors defined
  at archipelago scale
  
Make a model with species richness as the response, and with a random effect for 
  Archipelago. Species richness could be modeled as discrete count data (e.g., 
  a negative binomial distribution), but we’ll finish covering GLMMs later; for 
  now you can use log(Richness+1) to get a pretty normal looking response.
```{r}
island$log_richness <- log(island$Species + 1)

hist(island$log_richness) # great, looks normal
```

```{r}
mod_rich_arch <- lmer(log_richness ~ (1|Archipelago), data = island)
summary(mod_rich_arch)
ranef(mod_rich_arch)
```

What proportion of the variation in species richness occurs at the archipelago 
  scale, and what proportion occurs within archipelagoes? Which archipelagoes 
  are particularly diverse, and which are depauperate?  
- 0.64 variance at archipelagoe level
- 0.82 variance within archipelagoes
- particularly diverse; Canaries, Balearic
- particularly depauperate; Phoenix, Iles Crozet

Now let’s think about the six predictors. Make some exploratory plots of the 
  effect of each variable on richness, plotted at the appropriate scale. You’ll 
  need to merge the datasets. Think about which predictors might need to be 
  transformed for use in a linear model.
  
merge df
```{r}
df <- left_join(arch, island, by = 'Archipelago')
```

exploratory plots for each var against log_richness
```{r}
par(mfrow = c(3, 2))
plot(df$number.islands, df$log_richness) # maybe transform
plot(df$distance, df$log_richness)
plot(df$age, df$log_richness)
plot(df$Area, df$log_richness) # transform
plot(df$Elev, df$log_richness) # transform
plot(df$Temp, df$log_richness)
```

transform Elev and Area - huge range for these vars w/ big outliers
- also can transform number.islands
```{r}
plot(log(df$Area + 1), df$log_richness) # looks good, clear positive relationship
plot(log(df$Elev + 1), df$log_richness) # looks good, clear positive relationship
plot(log(df$number.islands + 1), df$log_richness) # more uniform, no clear relationship
```

predictor variable transformation in df
```{r}
df$log_area <- log(df$Area + 1)
df$log_elev <- log(df$Elev + 1)
df$log_islands <- log(df$number.islands + 1)

hist(df$log_area) # very normal dist.
hist(df$log_elev) # not perfect normal dist. but prob okay?
hist(df$log_islands) # now close to normal dist.
```

Construct mixed model(s) that test the roles of the six predictors in explaining 
  species richness. Plot fitted (fixed) effects as well as random effect estimates, 
  plus model diagnostics. 
  
```{r}
mod_rich_full <- lmer(log_richness ~ log_area + log_elev + log_islands +
                        distance + age + Temp +
                        (1|Archipelago), data = df)
summary(mod_rich_full)
```

fixed effects
```{r}
plot_model(mod_rich_full, type = 'est')
# estimate for log area, log elev, temp doesn't overlap 0
```

plot random effects
```{r}
plot_model(mod_rich_full, type = 're')
```

model diagnostics
```{r}
sim_res <- simulateResiduals(mod_rich_full)
plot(sim_res)
# observed vs. expected plot looks good
# quantile deviations detected, adjusted quantile test significant

plot_model(mod_rich_full, type = 'diag')
# no clear scary patterns here, residual distribution almost perfectly normal
```

Calculate how much variation the predictors explain, at the two different scales 
  in the data (island and archipelago)? I.e., present R2 values for the two scales. 
  Also, how much of the total variation have they explained, according to R2GLMM(m)?
- conditional R2 of .85 means the full model (w/ re) explained 85% of variation
- marginal R2 of .55 means the fixed effects (at island scale) explained 55% of
  variation
```{r}
r2_values <- r.squaredGLMM(mod_rich_full)

print(r2_values)
```

Use appropriate hypothesis tests (as described in lecture) to test the role of 
  the different predictors. 
```{r}
anova(mod_rich_full)
# no idea why I'm not getting p-values for this...

summary(mod_rich_full)
```
  
How do you interpret the results of these tests, and 
  the effects plots, in light of hypotheses for what controls species richness 
  in islands and archipelagos? What are the denominator degrees of freedom for 
  each predictor? This is essentially telling you how much replication there is 
  for that predictor, minus the number of parameters for that predictor. Do the
  denominator df make sense? Why or why not?
- I have no idea why anova(mod_rich_full) is not displaying p-values for me...
- but, after looking at the effects plots, I suspect that log area, log elevation,
  and temperature all have significant, positive effects on log species richness.
- This is supported by the massive f-value for log_area, and the large f-value
  for temperature. Perhaps log elevation is not actually a significant predictor,
  judging by it's pedestrian F-value. Really wish I had those p-values and the 
  time to troubleshoot this code!
- there's a bug somewhere in my code (perhaps b/c I recently updated R?) preventing
  me from seeing the denominator degrees of freedom. Wish I could tell you!

Is the model we’ve used the best model? Often I just stick with one big model 
  when the ratio of data to parameters is pretty good. But one might prefer to 
  find the best model(s), while accounting for model selection uncertainty. Use 
  AICc in some capacity to assess which predictors are important, what the ‘best’ 
  model is, and how sure you are about what the best model is. The details of 
  how you do it are up to you, as long as it seems justifiable. Remember to do 
  REML=FALSE for comparing models.
```{r}
# full
mod_rich_full <- lmer(log_richness ~ log_area + log_elev + log_islands +
                        distance + age + Temp +
                        (1|Archipelago), data = df, REML = FALSE)

mod_rich_no_area <- lmer(log_richness ~ log_elev + log_islands +
                        distance + age + Temp +
                        (1|Archipelago), data = df, REML = FALSE)

mod_rich_no_elev <- lmer(log_richness ~ log_area + log_islands +
                        distance + age + Temp +
                        (1|Archipelago), data = df, REML = FALSE)

mod_rich_no_temp <- lmer(log_richness ~ log_area + log_elev + log_islands +
                        distance + age +
                        (1|Archipelago), data = df, REML = FALSE)

mod_rich_no_dist <- lmer(log_richness ~ log_area + log_elev + log_islands +
                        age + Temp +
                        (1|Archipelago), data = df, REML = FALSE)

mod_rich_no_age <- lmer(log_richness ~ log_area + log_elev + log_islands +
                        distance + Temp +
                        (1|Archipelago), data = df, REML = FALSE)

mod_rich_no_island <- lmer(log_richness ~ log_area + log_elev +
                        distance + age + Temp +
                        (1|Archipelago), data = df, REML = FALSE)

arrange(AICc(
  mod_rich_full, mod_rich_no_area, mod_rich_no_elev, mod_rich_no_temp,
  mod_rich_no_dist, mod_rich_no_age, mod_rich_no_island), AICc)
```

Probably fine to stick with the full model here - essentially no difference in AIC
  between the full model, the model without island, or the model without age.
  You could remove these but it doesn't make a huge difference so probably
  best to just retain these predictors to avoid any potential headaches with 
  justifying their removal.
  