---
title: "hwk_13"
author: "Holden Jones"
date: "2025-05-14"
output: html_document
---


```{r}
library(tidyverse)
library(lme4)
library(ggeffects)
library(sjPlot)

ohia <- read_csv('data/ohia_merged.csv')
```


#1.


Using the seedlings that were harvested after the press drought treatment, 
  analyze how seedling height varies among treatments and populations. 
  
subset only harvested seedlings
```{r}
harvest <- ohia %>%
  filter(Fate == "harvest")
```

Be sure to include Block in your models, and think about which terms in the model 
  should be random effects, which should be fixed effects, which should be 
  fixed:random interactions, and which should be fixed:fixed interactions 
  (you will need all four).
- height as response variable
- block definitely as a random effect
- drought definitely as fixed effect
- population definitely as fixed effect
- drought and population as interaction
  
first, look at dist. of response variable
```{r}
hist(harvest$Height) # close to normal, long tail
```

Specifically, set up models to test whether populations differ in height, whether 
  the press drought treatment affects height, and whether populations differ in 
  how press drought affects height.
```{r}
mod_height_1 <- lmer(Height ~ Drought + 
                     (Drought | Population) + 
                     (1 | Block), 
                   data = harvest
                   )
summary(mod_height_1)
```
  
Make appropriate plots of model results and describe the magnitude of any 
  apparent relationships.
```{r}
plot(ggpredict(mod_height_1, terms = c("Drought"))) 
# drought has no impact on height
plot_model(mod_height_1, type = "re") 
# population v small impact, Poa, Kua diff
# block has no impact on height - this is good!
```

Next, set up models to test whether historical rainfall at a site predicts the 
  height of seedlings, and whether historical rainfall at a site predicts the 
  effect of press drought on height. 
```{r}
mod_height_2 <- lmer(Height ~ MAR * Drought + 
                    (1|Population) + (1 | Block), 
                    data = harvest)
summary(mod_height_2)
```

Make appropriate plots of model results and describe the magnitude of any 
  apparent relationships.
```{r}
plot(ggpredict(mod_height_2, terms = c("MAR", "Drought")))
# no correlation between historical rainfall and height
# no sig difference in this relationship for control vs. drought
plot_model(mod_height_2, type = "re")
# only Poamoho 2 sig higher height, not by much
# block I barely sig taller than the others - probably not meaningful
```
  
Finally, set up models to test whether historical evapotranspiration at a site 
  predicts the height of seedlings, and whether historical evapotranspiration at 
  a site predicts the effect of press drought on height. 
```{r}
mod_height_3 <- lmer(Height ~ MET * Drought + 
                    (1|Population) + (1 | Block), 
                    data = harvest)
summary(mod_height_3)
```
  
Make appropriate plots of model results and describe the magnitude of any 
  apparent relationships.
```{r}
plot(ggpredict(mod_height_3, terms = c("MET", "Drought")))
# no correlation between evapotranspiration and height
# no sig difference in this relationship for control vs. drought
plot_model(mod_height_3, type = "re")
# no random effect difference for population
# no difference for block here
```
  
How do you interpret the results?
- drought treatment has no significant impact on height
- Neither evapotranspiration or rainfall have a significant effect on height,
  regardless of drought treatment
- There are some differences between populations in height, notably Kuakola and 
  Puamoho 2 populations producing significantly taller seedlings
- Block likely has no impact on height, which is good (note small sig higher
  for Block 1 in the 2nd model)


#2.

Using the seedlings that were fated to be harvested after the press drought 
  treatment, analyze how survival during the press drought is affected by the 
  various potential predictors. 
  
Be sure to think about what the appropriate probability distribution is for the 
  response variable. 
- survival as response variable going to be binomial dist.
```{r}
hist(harvest$Dead)
```

Specifically, set up models to test whether populations vary in probability of 
  mortality (i.e., the response column DEAD), whether the press drought treatment 
  affects mean mortality, and whether populations vary in how mortality is 
  affected by the press drought treatment. 
```{r}
mod_dead_1 <- glmer(Dead ~ Drought + (Drought | Population) + (1| Block), 
                 data = harvest, 
                 family = binomial)
summary(mod_dead_1) # sig effect of drought
```

effects plots
```{r}
plot(ggpredict(mod_dead_1, terms = "Drought"))
# higher mortality in drought treatment
plot_model(mod_dead_1, type = "re")
# doesn't seem like any difference in effect of drought by population
# does seem to be some differences in mortality across blocks - uh oh
```

Next, set up models to test whether historical rainfall at a site predicts 
  mortality, and whether historical rainfall at a site predicts the effect of 
  press drought on mortality. 
```{r}
mod_dead_2 <- glmer(Dead ~ MAR*Drought + (1 | Population) + (1 | Block), 
                 data = harvest, 
                 family = binomial)
summary(mod_dead_2) # model nearly unidentifiable: very large eigenvalue
# MAR has sig positive effect on death
```

```{r}
plot(ggpredict(mod_dead_2, terms = c("MAR", "Drought")))
# increased MAR increases the mortality of seedlings, regardless of treatment
# but this happens equally for drought vs. control treatment
plot_model(mod_dead_2, type = "re")
# no difference across populations, suspiciously so?
# again, does seem like some differences in block for mortality - suspicious
```
  
Finally, set up models to test whether historical evapotranspiration at a site 
  predicts mortality, and whether historical evapotranspiration at a site predicts 
  the effect of press drought on mortality. 
```{r}
mod_dead_3 <- glmer(Dead ~ MET * Drought + (1 | Population) + (1 | Block), 
                 data = harvest, 
                 family = binomial)
summary(mod_dead_3) # model failed to converge?
```

```{r}
plot(ggpredict(mod_dead_3, terms = c("MET", "Drought")))
# does seem that increased MAT leads to increased mortality, and that the slope
# of this relationship is different between control and drought treatments, but
# this relationship is not statistically significant
plot_model(mod_dead_3, type = "re")
# no difference across populations, but yes sig difference in mortality across blocks
```

How do you interpret the results?
- drought treatment increases mortality vs. control treatment
- higher rainfall at source increases mortality, equally for control / drought
- higher evapotranspiration at source does not impact mortality
- no difference in mortality across source populations
- see some impact of blocks in the greenhouse on mortality - not good!

#3.

Using the seedlings that were fated for the terminal drought experiment 
  (i.e., the seedlings that were not harvested), analyze how longevity during 
  terminal drought is affected by the various potential predictors. 
  
subset for terminal drought seedlings
```{r}
drought <- ohia %>%
  filter(Fate == "terminaldrought")
```
  
Specifically, set up models to test whether populations vary in longevity, 
  whether the press drought treatment affects longevity, and whether populations 
  vary in how longevity is affected by the press drought treatment. 
```{r}
mod_long_1 <- lmer(Longevity ~ Drought + (Drought | Population) + (1| Block), 
               data = drought)
summary(mod_long_1)
```

```{r}
plot(ggpredict(mod_long_1, terms = c("Drought")))
# no differences in longevity in control vs. drought treatments
plot_model(mod_long_1, type = "re")
# no big difference in longevity for different populations
# looks like D survived less time, A survived longer for block effects
```

Next, set up models to test whether historical rainfall at a site predicts 
  longevity, and whether historical rainfall at a site predicts the effect of 
  press drought on longevity.
```{r}
mod_long_2 <- lmer(Longevity ~ MAR * Drought + (1 | Population) + (1 | Block), 
                   data = drought)
summary(mod_long_2)
```

```{r}
plot(ggpredict(mod_long_2, terms = c("MAR", "Drought")))
# increasing rainfall of source population decreases longevity of seedlings
plot_model(mod_long_2, type = "re")
# no big difference in longevity for different populations
# looks like D survived less time, A survived longer for block effects
```

Finally, set up models to test whether historical evapotranspiration at a site 
  predicts longevity, and whether historical evapotranspiration at a site predicts 
  the effect of press drought on longevity. 
```{r}
mod_long_3 <- lmer(Longevity ~ MET * Drought + (1|Population) + (1 | Block), 
                   data = drought)
summary(mod_long_3)
```

```{r}
plot(ggpredict(mod_long_3, terms = c("MET", "Drought")))
# increasing evapotransporation of source population decreases longevity of seedlings
# this relationship appears same for both drought treatments
plot_model(mod_long_3, type = "re")
# big difference in longevity for different populations - PTAB, Manana1 high
# no difference in block as random effect - why does this look diff from last model?
```

How do you interpret the results?
- no differences in longevity for drought vs. control treatments
- no big differences in longevity for different source populations
- looks like block D survived less than others, while block A survived more
- MAR appears to be driving longevity, increasing rainfall of source population
  leads to decreased survival longevity in terminal drought experiment
- MET also appears to be driving longevity (or at least correlated), increasing
  evapotranspiration of source population leads to decreased survival longevity
  in terminal drought experiment.



