---
title: "hwk_6"
author: "Holden Jones"
date: "2025-04-10"
output: html_document
---

```{r}
library(tidyverse)
library(arm)
library(ggeffects)
library(DHARMa)
library(car)
library(ggResidpanel)

birds <- read_csv('data/pox_data.csv')
```

#1. 

number of each species at each elevation
- not surprisingly little to no APAP and IIWI at low and mid elevations
```{r}
# factor species and elevation for plotting
birds$Elev <- factor(birds$Elev, levels = c("low", "mid", "high"))
birds$Species <- factor(birds$Species)

birds_spe_ele <- birds %>%
  group_by(Species, Elev) %>%
  count()

birds_spe_ele
```

prevalence of malaria for each species at each elevation (only 0 or 1)
```{r}
# filter out ind. where malaria = 2
birds_mal <- birds %>%
  filter(Malaria %in% c(0, 1))

birds_mal # now have 7,538 birds whose malaria status we know

birds_mal %>%
  group_by(Species, Elev, Malaria) %>%
  count()
```


#2.

first look at hist of data
- many more 0s than 1s
```{r}
discrete.histogram(birds$Activepox)
```

model testing whether pox prevalence depends on elevation and between species and 
whether different species depends on different elevations
- this is binomial data, use binomial dist.
- most interested in modeling p, probability of pox
```{r}
glm_pox_bin <- glm(Activepox ~ Species + Elev + Species*Elev, data = birds,
                   family = binomial)
summary(glm_pox_bin)
```

evaluate model using general methods in this course
- Iiwi has sig higher pox prevalence than Apapane
- both low and mid elevations have sig higher pox prevalence than high elevation
- compared to Apapane at high elevations (I'm not sure why additional comparisons are not shown); 
  - Amakihi low sig lower
  - White eye low sig lower
  - Amakihi mid sig lower
  - White eye mid sig lower

NA for Iiwi low b/c this species is not present at low elevations

likelihood ratio tests
- each of these predictors has a significant effect on pox prevalence
```{r}
Anova(glm_pox_bin)
```

look at residual, effects plots
```{r}
sim_res_model <- simulateResiduals(fittedModel = glm_pox_bin)
plot(sim_res_model)

resid_panel(glm_pox_bin, plots = 'resid', smoother = TRUE)

plot(ggpredict(glm_pox_bin, terms = c("Elev", "Species")))
```

All residuals look great, I don't see any issues with overdispersion - binomial
  distribution appears to be an appropriate choice here.
  
The effects plot is especially helpful
- Amakihi and white eye appear resilient to pox across elevations
- almost no pox at high elevations
- Apapane and Iiwi highly susceptible at low and mid elevations
- notice that error bars are super high at low elevations, b/c not many of these
  native species


#3.

filter dataset for only birds with malaria status, no Iiwi
```{r}
# use birds_mal from above, filter out Iiwi
birds_mal <- birds_mal %>%
  filter(Species != 'IIWI')
```

create model testing if malaria status and elevation affect pox prevalence
- include species interactions for each predictor
```{r}
glm_pox_mal <- glm(Activepox ~ Malaria + Elev + Species*Elev + Species*Malaria, 
                   data = birds_mal,
                   family = binomial)

summary(glm_pox_mal)
```

interpreting model summary
- malaria has a significant positive effect on pox prevalence
- high elevations have a significantly lower pox prevalence than low and mid
- interactions show species dependent effects
  - Amakihi are sig less likely to have pox if they have malaria than other species(?)
  - Amakihi are sig more likely to have pox at high elevations compared to other species(?)

evaluate model
- each of these predictors and the interactions have significant effects
```{r}
Anova(glm_pox_mal)
```

look at residual, effects plots
- no issues with residuals! binomial dist. fits well here
- Warbling white eye not included here (likely didn't collect malaria data b/c resistent)
- effects plots shows what we would expect
  - lots of uncertainty for pox probability at low elevations for apapane (small sample)
  - Apapane are more likely to have pox at mid elevations than Amakihi
- at high elevation see Amakihi have small prob of pox but higher than Apapane
- malaria greatly increases probability of pox for apapane, minimally for amakihi
  - it does so at both mid and high elevations, not low (small sample size of native birds)
```{r}
sim_res_model <- simulateResiduals(fittedModel = glm_pox_mal)
plot(sim_res_model)

resid_panel(glm_pox_mal, plots = 'resid', smoother = TRUE)

plot(ggpredict(glm_pox_mal, terms = c("Elev", "Species")))

plot(ggpredict(glm_pox_mal, terms = c("Malaria", "Species")))

plot(ggpredict(glm_pox_mal, terms = c("Malaria", "Elev")))
```


