---
title: "hwk_5"
author: "Holden Jones"
date: "2025-04-08"
output: html_document
---

```{r}
library(tidyverse)
library(GGally)
library(effects)
library(MASS)

fish <- read_csv('data/CRCP_fish.csv')
```


#1. 
look at dist of ta(20), hard_coral (17), sand(21), cca(22)
- this function is super cool!
```{r}
compare_var <- ggpairs(fish, columns = c(17,20,21,22))
compare_var
```


#2.
single predictor models for each predictor with kole count as response
- use Poisson dist.
- which predictors should be transformed to reduce skew?
  - based on dist. looks like hard_coral, cca, sand are very skewed
```{r}
fish <- fish %>%
  mutate(sqrt_hard_coral = sqrt(hard_coral),
         sqrt_sand = sqrt(sand),
         sqrt_cca = sqrt(cca))
```
  
check out these sqrt vars
- doesn't this make the vars closer to the normal dist? but it does reduce skew
```{r}
compare_sqrt_var <- ggpairs(fish, columns = 48:50)
compare_sqrt_var
```

positive effect on kole count
-hard_coral, cca

negative effect on kole count
- sand, ta
```{r}
# hard_coral
glm_hard_coral <- glm(count ~ hard_coral, family = poisson, data = fish)
summary(glm_hard_coral) # hard_coral sig pos effect

glm_hard_coral_sqrt <- glm(count ~ sqrt_hard_coral, family = poisson,
                           data = fish)
summary(glm_hard_coral_sqrt) # sqrt hard_coral sig pos effect

# sand
glm_sand <- glm(count ~ sand, family = poisson, data = fish)
summary(glm_sand) # sand sig neg effect

glm_sand_sqrt <- glm(count ~ sqrt_sand, family = poisson,
                           data = fish)
summary(glm_sand_sqrt) # sqrt sand sig neg effect

# cca
glm_cca <- glm(count ~ cca, family = poisson, data = fish)
summary(glm_cca) # cca sig pos effect

glm_cca_sqrt <- glm(count ~ sqrt_cca, family = poisson,
                           data = fish)
summary(glm_cca_sqrt) # sqrt cca sig pos effect

# ta - not transformed
glm_ta <- glm(count ~ ta, family = poisson, data = fish)
summary(glm_ta) # ta sig neg effect
```

let's look at overdispersion
- every single model is extremely overdispersed
- ideally dispersion would be 1 - most of these are about 30! even the transformed
  ones
```{r}
# make function for calculating dispersion
dispersion = function(model) {
  sum(residuals(model, type = "pearson")^2) / df.residual(model)
}

# calculate degree of dispersion for each model
dispersion(glm_hard_coral)
dispersion(glm_hard_coral_sqrt)
dispersion(glm_sand)
dispersion(glm_sand_sqrt)
dispersion(glm_ta)
dispersion(glm_cca)
dispersion(glm_cca_sqrt)
```

likelihood ratio tests for each model
- the single predictor is significant in every single one of these likelihood
  ratio tests
- I don't like that I can't see the sign or size of the effect
```{r}
anova(glm_hard_coral)
anova(glm_hard_coral_sqrt)
anova(glm_sand)
anova(glm_sand_sqrt)
anova(glm_ta)
anova(glm_cca)
anova(glm_cca_sqrt)
```

make effects plots to investigate size and sign of effects
- hard_coral has positive effect on count
- sand has negative effect on count
- ta has negative effect on count
- cca has positive effect on count
```{r}
plot(allEffects(glm_hard_coral))
plot(allEffects(glm_hard_coral_sqrt))
plot(allEffects(glm_sand))
plot(allEffects(glm_sand_sqrt))
plot(allEffects(glm_ta))
plot(allEffects(glm_cca))
plot(allEffects(glm_cca_sqrt))
```

All in all, it appears that hard coral and sand have significant effects on count,
  and that ta and cca have negative effects on count. BUT - I'm spooked by the 
  massive amount of dispersion in every model (even those that I transformed).
  I don't feel that these models fit well enough for me to appropriately interpret.
  

#3.
use negative binomial distribution when Poisson dist. is overdispersed
- re run every model changing family to negative binomial
- need to use glm.nb() from MASS package
```{r}
# hard_coral
glm_hard_coral <- glm.nb(count ~ hard_coral, data = fish)
summary(glm_hard_coral) # hard_coral sig pos effect

glm_hard_coral_sqrt <- glm.nb(count ~ sqrt_hard_coral, data = fish)
summary(glm_hard_coral_sqrt) # sqrt hard_coral sig pos effect

# sand
glm_sand <- glm.nb(count ~ sand, data = fish)
summary(glm_sand) # sand sig neg effect

glm_sand_sqrt <- glm.nb(count ~ sqrt_sand, data = fish)
summary(glm_sand_sqrt) # sqrt sand sig neg effect

# cca
glm_cca <- glm.nb(count ~ cca, data = fish)
summary(glm_cca) # cca sig pos effect

glm_cca_sqrt <- glm.nb(count ~ sqrt_cca, data = fish)
summary(glm_cca_sqrt) # sqrt cca sig pos effect

# ta - not transformed
glm_ta <- glm.nb(count ~ ta, data = fish)
summary(glm_ta) # ta sig neg effect
```

calculate degree of dispersion for these new neg binomial models
- that's more like it! this is now appropriate for interpretation
- dispersion parameter approximately equal to 1 (way better than 30)
- the negative binomial fits the data much better than the Poisson dist.
```{r}
dispersion(glm_hard_coral)
dispersion(glm_hard_coral_sqrt)
dispersion(glm_sand)
dispersion(glm_sand_sqrt)
dispersion(glm_ta)
dispersion(glm_cca)
dispersion(glm_cca_sqrt)
```

likelihood ratio tests for each of these negative binomial models as well
```{r}
anova(glm_hard_coral)
anova(glm_hard_coral_sqrt)
anova(glm_sand)
anova(glm_sand_sqrt)
anova(glm_ta)
anova(glm_cca)
anova(glm_cca_sqrt)
```

look at effects plots for new models
```{r}
plot(allEffects(glm_hard_coral))
plot(allEffects(glm_hard_coral_sqrt))
plot(allEffects(glm_sand))
plot(allEffects(glm_sand_sqrt))
plot(allEffects(glm_ta))
plot(allEffects(glm_cca))
plot(allEffects(glm_cca_sqrt))
```

This new round of negative binomial models fits the data much better.
- These negative binomial models are no longer overdispersed
- The overall findings are still the same (hard_coral, cca sig pos; sand, ta sig neg)
- It appears that square root transforming skewed variables is no longer necessary
  minimal difference in model fit between transformed and non-transformed models
  

#4.

make full model with multiple substrate types
- not using transformed vars per earlier observation
```{r}
glm_full <- glm.nb(count ~ hard_coral + sand + cca + ta, data = fish)
summary(glm_full) # hard_coral sig pos, cca sig pos, ta sig pos
```

likelihood ratio tests and effects plots
```{r}
dispersion(glm_full)
anova(glm_full)
plot(allEffects(glm_full))
```

overall conclusions
- with the full model it now appears that all substrates have a positive effect
  on counts? This does seem a bit unlikely.
- this full model is not overdispersed (although it's close to limit)
- I don't feel the need to run any more analyses, time for dinner!

