---
title: "hwk_1"
author: "Holden Jones"
date: "2025-01-13"
output: html_document
---

setup
```{r}
library(tidyverse)
library(effects)

bird_data <- read_csv('data/foraging.height.edit.csv')
```

# 1. 
Start with some data exploration. 

How many observations are there of each bird species in each experimental
treatment? 
```{r}
bird_data %>%
  group_by(SPECIES, Rat_Removal) %>%
  summarize(count = n())
```

Make a boxplot or violinplot to show the distribution of foraging heights of each bird species in each experimental treatment. 
```{r}
ggplot(bird_data, aes(Rat_Removal, foraging.ht.m, 
                 fill = SPECIES)) +
  geom_violin()
```

What have you learned so far?

- Lots of variation in the number of observations for each species, but they 
  are closely split 50/50 across treated / untreated sites.
- Seems to be more variation in foraging height for most species in the treated
  sites. They forage at lower heights in the treated sites.

# 2.

Now make a plot showing the mean foraging height of each bird species in each
treatment, and include error bars displaying +/- one standard error of the mean.
```{r}
ggplot(bird_data, aes(Rat_Removal, foraging.ht.m, 
                 fill = SPECIES)) +
  geom_boxplot()
```

What is the meaning of the standard error of the mean? How does this plot differ
from the plot in #1? 

- Standard Error of the Mean incorporates variation within a sample and sample
  size to show how likely the observed sample mean is to be within the true
  population mean
- This plot incorporates standard error with error bars which contain the true
  mean with 95% confidence

# 3.
Fit a linear model that tests whether the bird species differ in foraging height,
whether the rat removal treatment affects foraging height, and whether the 
effect of the rat removal differs between the bird species. 
```{r}
height_model <- lm(foraging.ht.m ~ SPECIES + Rat_Removal + Rat_Removal*SPECIES , 
            data = bird_data)
summary(height_model)
```

Make plots to assess the distribution of the residuals. 
```{r}
par(mfrow = c(2, 2))
plot(height_model)
```

Report F-tests for the terms of the model. 
```{r}
anova(height_model)
```

- SPECIES, 3.50, 0.004
- Rat_Removal, 16.84, <0.001
- SPECIES*Rat_Removal, 0.42, 0.84

Create an effects plot displaying the fitted effects. 
```{r}
plot(allEffects(height_model))
```

How do you interpret the results? What are the magnitudes of the effects? 

- The magnitude of the effects is pretty minimal. HAEL decreases foraging height
  by 3.06m in untreated  vs. treated sites, while JAWE increases by 2.44m in 
  untreated vs. treated sites.
- Overall interpretation; removing rats has an impact on the foraging height of
  birds, although some are more affected than others. In general, birds tend to
  forage at lower heights when rats are removed.
  
# 4.
So far we have examined the effect of the rat removal, but we have not considered
the role of kipuka size, or the role of food availability (which is influenced in
part by kipuka size). Make a new model that adds kipuka area and arthopod biomass
as predictors, and include interactions that test whether the treatment effect
depends on these additional predictors.
```{r}
super_model <- lm(foraging.ht.m ~ SPECIES + 
                    Rat_Removal + 
                    Rat_Removal*SPECIES +
                    Area_ha +
                    tot.arth.bm +
                    Rat_Removal*Area_ha +
                    Rat_Removal*tot.arth.bm, 
            data = bird_data)
summary(super_model)
```

Kipuka area and arthopod biomass are both continuous predictors – before you add
them to the model, make some scatterplots to assess whether these predictors
should be transformed when including them in the model. When assessing predictors
for transformation, it doesn’t matter if the predictors are normally distributed
(this only matters for the response variable) – what matters is whether the
predictors are very skewed, such that a few outlying points will have a large
influence on a fitted regression line. For skewed predictors, a log or square root
transformation will generally help.
```{r}
ggplot(bird_data, aes(x = Area_ha, y = tot.arth.bm)) +
  geom_point()

# also can use individual histograms:
ggplot(bird_data, aes(x = Area_ha)) + # bimodal, not necessarily skewed
  geom_histogram(binwidth = 1)

ggplot(bird_data, aes(x = tot.arth.bm)) + # very skewed
  geom_histogram(binwidth = 1)
```

- tot.arth.bm looks very skewed (one point)
```{r}
# log transform tot.arth.bm
bird_data <- bird_data %>%
  mutate(log_tot.arth.bm = log(tot.arth.bm + 1))  # Adding 1 to avoid log(0)

# cool, not skewed anymore!
ggplot(bird_data, aes(x = log_tot.arth.bm)) + # very skewed
  geom_histogram(binwidth = 1)
```

Report F-tests and effects plots. 
```{r}
log_super_model <- lm(foraging.ht.m ~ SPECIES + 
                    Rat_Removal + 
                    Rat_Removal*SPECIES +
                    Area_ha +
                    log_tot.arth.bm +
                    Rat_Removal*Area_ha +
                    Rat_Removal*log_tot.arth.bm, 
            data = bird_data)
summary(log_super_model)

anova(log_super_model)
```

- f-tests for each term of model above - effects plots below
- SPECIES, Rat_Removal, Area_ha, log_tot.arth.bm all highly sig terms
```{r}
plot(allEffects(log_super_model))
```

How do the results of this model differ from the model in #3? 

- SPECIES and Rat_Removal are still significant terms, but we also now know that
  area and log arthropod biomass have large significant effects on foraging height
- Also the interaction between SPECIES and Rat_Removal is no longer a sig term

How do you interpret the results at this stage?

- The presence of rats, the bird species, the area of the kipuka, and the biomass 
  of arthropods all have significant effects on foraging height, but that none of
  the interactions between these terms are significant.
  
# 5. 
One thing we have not accounted for in our models is the fact that the study design 
involves repeated measurements of the same kipuka (and potentially the same birds) 
over time. We will learn how to properly account for repeated measures, by adding 
a random effect for kipuka, later in the course. However, there is a different 
additional predictor that may be helpful in testing the treatment effect. Add 
DATE as a factor to the linear model from #4. 
```{r}
log_super_date_model <- lm(foraging.ht.m ~ SPECIES + 
                    Rat_Removal + 
                    Rat_Removal*SPECIES +
                    Area_ha +
                    log_tot.arth.bm +
                    DATE +
                    Rat_Removal*Area_ha +
                    Rat_Removal*log_tot.arth.bm, 
            data = bird_data)
summary(log_super_date_model)

plot(allEffects(log_super_date_model))

anova(log_super_date_model)
```

How does the inclusion of the date of sampling alter the model results? 

- All sig effects from previous model are still sig
- Date is now sig as well, as is interaction of Rat_Removal and log_tot.arth.bm

Why do you think that is?

- tons of variation across the different dates. There are 27 unique dates, so 
  of course there's gonna be some with lower height foraging. But that doesn't
  really tell us much. It would be perhaps more interesting to divide by season
  or look at month (ie so we could compare across different years). Don't really
  see a clear trend in the different dates. I'm not sure why this now makes the 
  interaction b/ween Rat_Removal and log_tot.arth.bm significant though.